Thesis: draft

### 2021-03-17 draft.tex
made some small changes

One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence of this is that if a network possesses a more skewed degree distribution, the SBM will put the nodes with a high degree in a small group, together with a high probability of connecting to other, larger groups \cite{Clauset2017}. The reason behind this is one of the main assumptions of the stochastic block model, namely the fact that the edge probabilities only depend  on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically  equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. However, a network with a skewed degree distribution violates this assumption of edge independence and the SBM seeks to explain this over other kinds of latent group structures that we might care about \cite{Clauset2017}. Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us to far astray, but interested readers can find more detailed information in reference \cite{Karrer2011}. 

@@

There are, however, some properties that are different between on-line and off-line networks as well. One of them might be the degree correlations/mixing pattern. It is commonly assumed and accepted that most real-life social networks have an assortative mixing pattern (this distinguishes them from other real-world networks, eg. biological networks \cite{F.Costa2007}, which often have a disassortative pattern) \cite{Hu2009}\cite{Zhang2014}. For a long time, it was thought that this is also valid for on-line social networks. However, recent research has shown that a lot of on-line social networks display disassortative mixture \cite{Hu2009}\cite{Zhang2014}. This discrepancy between real-life social networks and on-line social networks may be explained as follows. In real life, `ordinary people' often want to be friends with celebrities, however, the celebrities usually prefer to stay within their own circles \cite{Hu2009}\cite{Zhang2014}. The `ordinary people' don't have access to these circles and we are left with the situation where `ordinary people' interact with `ordinary people' and celebrities hang out among other celebrities; hence, an assortative structure. In on-line social networks, `ordinary people' can easily connect with the celebrity and vice-versa, the celebrity wants to show his/her importance/influence by the number of fans \cite{Hu2009}\cite{Zhang2014}. This leads to a disassortative mixing pattern.

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. The current chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
\newline
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary/discrete models and continuous approaches. These will be discussed in the following two sections, Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn people is discussed and, in Section \ref{thesisOp}, a summary of the opinion dynamics model used in this thesis is given.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent is assigned an opinion; then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive the system to converge to unanimity; and, does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? \\
Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan \cite{Benjamini2016} (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of notion of `minority' and `majority' opinion). It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}.
% https://arxiv.org/pdf/0710.3256.pdf Contains a mean field mathimatical description, interesting?

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will updates his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

\subsection{The classical models}

@@

In bounded confidence models, each node has, beside an opinion, a threshold $\epsilon$, also called an uncertainty or tolerance (note that $\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are then ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. The two most popular bounded confidence models are the Deffuant model and the Hegselmann-Krause model \cite{Castellano2009}. \\

@@

The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (so the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinion inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends in the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that human beings can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{Modelling opinion dynamics in the age of algorithmic personalization} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $Q$ elements
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC and REF methods are mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

@@

It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure doesn't change (in the sense that each user/node has a fixed number of neighbors that doesn't change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which aren't. Hence, in this sense, this does affect the network structure in some way, because if a user, for example, doesn't become active for a long period of time, he/she doesn't participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and different things like connectivity etc. \\

@@

In this thesis we follow the same activation mechanism as was proposed in the paper of prof. Nicola Perra and prof. Luis E.C. Rocha (see reference \cite{Perra2019}). Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Prof. Nicola Perra and prof. Luis E.C. Rocha choose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\

@@

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that the other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note here that being stubborn and being extreme doesn't need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is a tendency that people with extreme opinions are more persistent towards these opinions than people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Turner}).\\ 
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of such extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore wouldn't reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge impacts. \\
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number, often between $\interval{0}{1}$ or between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is again based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it. 
% talk about results here??

@@

	\item If $r_i = 1$, the node is completely stubborn and there is no change of convincing him/here of the other idea

@@

@@

\textbf{The threshold model}

@@

In the threshold model one introduces a notion of stubbornness by introducing a threshold parameter $T$. This parameter can be either the same for all the nodes ($T$) or can be made individual ($T_i$). The main idea is then that nodes can only change their opinion to the other opinion if the fraction of friends with that opinion ($p$) is bigger than $T$ ($p > T$). The idea behind this threshold model is that people are a bit hesitant to change their opinion, but if they have enough friends with another opinion they might be convinced of that opinion as well. The more persistent you are, the more friends you need in order to be convinced of another idea. \\
There are different possibilities to implement this kind of model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_A$ and $p_B$ respectively, is determined; then the following rules are used to update the opinion

@@

% Talk about probabilistic implementation?

@@

\chapter{Network models: Derivations}
\section{The Erd\H{o}s-R\'{e}nyi model}\label{ER}
\subsection{Degree distribution}\label{degdis}
The probability that a node is connected to $k$ other nodes and not to $N-1-k$ others is given by
\begin{equation}
	p^k (1-p)^{N-1-k}
\end{equation}
The number of ways to choose the $k$ nodes among the $N-1$ possible candidates is given by the binomial coefficient
\begin{equation}
	\binom{N-1}{k} = \frac{(N-1)!}{k!(N-1-k)!}
\end{equation}
and the probability of being connected to exactly $k$ other nodes becomes
\begin{equation}\label{degDistRan}
	p_k = \binom{N-1}{k}p^k (1-p)^{N-1-k}
\end{equation}
We thus derived the degree distribution of a random network.

>>>>

One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence of this is that if a network possesses a more skewed degree distribution, the SBM will put the nodes with a high degree in a small group, together with a high probability of connecting to other, larger groups \cite{Clauset2017}. The reason behind this is one of the main assumptions of the stochastic block model, namely the fact that the edge probabilities only depend  on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically  equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. However, a network with a skewed degree distribution violates this assumption of edge independence and the SBM seeks to explain this over other kinds of latent group structures that we might care about \cite{Clauset2017}. Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us too far astray, but interested readers can find more detailed information in reference \cite{Karrer2011}. 

@@

There are, however, some properties that are different between on-line and off-line social networks as well. One of them might be the degree correlations/mixing pattern. It is commonly assumed and accepted that most real-life social networks have an assortative mixing pattern (this distinguishes them from other real-world networks, eg. biological networks \cite{F.Costa2007}, which often have a disassortative pattern) \cite{Hu2009}\cite{Zhang2014}. For a long time, it was thought that this is also valid for on-line social networks. However, recent research has shown that a lot of on-line social networks display disassortative mixture \cite{Hu2009}\cite{Zhang2014}. This discrepancy between real-life social networks and on-line social networks may be explained as follows. In real life, `ordinary people' often want to be friends with celebrities, however, the celebrities usually prefer to stay within their own circles \cite{Hu2009}\cite{Zhang2014}. The `ordinary people' don't have access to these circles and we are left with the situation where `ordinary people' interact with `ordinary people' and celebrities hang out among other celebrities; hence, an assortative structure. In on-line social networks, `ordinary people' can easily connect with the celebrity and vice-versa, the celebrity wants to show his/her importance/influence by the number of fans \cite{Hu2009}\cite{Zhang2014}. This leads to a disassortative mixing pattern.

@@

\chapter{Opinion dynamics}
\label{chap3}
The study of opinion dynamics in a group of people, as already mentioned in the introduction, has two major layers: a network layer and an opinion dynamics layer. The network layer, which describes the underlying structure and correlations between the agents/people, is intensively described in the previous chapter, Chapter \ref{chap2}. The current chapter focuses on the second important layer, namely that of the opinion dynamics models. These models need to describe, in a simplified way, how opinions are formed and how they evolve in a group of people that interact in a certain way. Hence, these opinion dynamics are the dynamics that run on the underlying network structure.\\
Several models have been proposed; here we will review some of the most important ones, with the main focus on the models used in this thesis.\\
\newline
Some major attention will also be given to the content curation/filtering algorithms used by social media companies (see Section \ref{filter}). This is needed because this thesis deals with opinion dynamics on on-line social media, so the aspect of filtering cannot be ignored in the formation and evolution of opinions.\\
\newline
There will also be some attention on the activation mechanism (see Section \ref{actMech}). This is the mechanism that describes the timings in which content/information is exchanged between users \cite{Perra2019}. Such a mechanism may change the network from a static one to a more temporal one, where nodes become active and inactive over time. \\
\newline
There are two major groups of opinion dynamics models, namely binary/discrete models and continuous approaches. These will be discussed in Section \ref{Bin} and Section \ref{Con}, respectively. This thesis only deals with binary opinion dynamics models, so these will be thoroughly described; the continuous models will be quickly reviewed in order to give a more complete overview.\\
\newline
Finally, in Section \ref{stubb}, the introduction of stubborn people is discussed and, in Section \ref{thesisOp}, a summary of the opinion dynamics model used in this thesis is given.

@@

One of the simplest opinion dynamics models is the binary majority model. In this model each agent has one of the two opinions A or B (or 0 or 1). Initially, these opinions can be equally distributed or not. If the distribution is not equal, but instead eg. 80/20, there is a predominant opinion and a minority opinion.\\
The following updating rule is then imposed on the network: each agent/individual adopts the majority opinion of his/her nearest neighbors \cite{Nguyen2020}. Thus, initially each agent is assigned an opinion; then the opinions are evolved in time: each individual iteratively updates his/her opinion according to the majority opinion of his/her direct neighbors \cite{Nguyen2020}.\\
\newline
Several studies have been performed on this kind of model, searching for answers on several questions such as: what are the conditions that drive the system to converge to unanimity; and, does an opinion, that is initially a majority opinion, remain a majority opinion in the final state \cite{Nguyen2020}? \\
Most of these studies conclude that, at the end of the opinion evolution, a complete consensus state is obtained where all the agents adopt the same opinion \cite{Nguyen2020}. This is even the case when the opinions are equally and randomly distributed in an Erd\H{o}s-R\'{e}nyi network (with the side note that the degree needs to be high enough), as was shown by Benjamini, Chan, O'Donnell, Tamuz and Tan \cite{Benjamini2016} (also note that even when the opinions are equally distributed, there may still be a slight imbalance due to the random assignement of an opinion to each node; instead of a perfect 50/50 distribution, one may end up with a eg. 49.8/50.2 distribution; in this sense there is still some kind of notion of `minority' and `majority' opinion). It is also found that the rate of convergence increases with increasing edge density and that if the network is sparse enough, consensus will not be reached and the two opinions coexist \cite{Nguyen2020}. This can be explained by the fact that if the network is sparse enough, we may have disconnected parts; each of these disconnected groups may then evolve to a different consensus opinion, so that over the entire network different opinions coexist \cite{Nguyen2020}.
% https://arxiv.org/pdf/0710.3256.pdf Contains a mean field mathimatical description, interesting?

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will update his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

\subsection{The classical models}\label{classMod}

@@

In bounded confidence models, each node has, beside an opinion, a threshold $\epsilon$, also called an uncertainty or tolerance (note that $\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. The two most popular bounded confidence models are the Deffuant model and the Hegselmann-Krause model \cite{Castellano2009}. \\

@@

The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (so the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinions inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends in the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{Modelling opinion dynamics in the age of algorithmic personalization} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $Q$ elements. This filtering method is based on the semantic filtering explained above
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC and REF methods are mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

@@

It might, however, be a more realistic situation if nodes/users become active/inactive over time. This would be a better representation of real social media platforms, where users are active for some periods of time and then become inactive again. This kind of activation mechanism of the users gives the network a more temporal character. The underlying network structure doesn't change (in the sense that each user/node has a fixed number of neighbors that doesn't change over time, so no new friends are made or, conversely, no old friends are being defriended), but by making users active and inactive the exchange of information between users becomes time dependent and depends on which users are active and which aren't. Hence, in this sense, this does affect the network structure in some way, because if a user, for example, doesn't become active for a long period of time, he/she doesn't participate in the exchange of information/opinions and the network structure behaves like if this user/node is absent. This may affect the dynamics, since it may seem as if the network has a different average degree and some properties like connectivity, clustering coefficient, etc may behave differently. \\

@@

In this thesis we follow the same activation mechanism as was proposed in the paper of prof. Nicola Perra and prof. Luis E.C. Rocha (see reference \cite{Perra2019}). Every time step users are made active with some probability $p_i$ (thus, every time step, each node becomes active with a probability $p_i$). In this thesis the activation probability is set equal to $p = 0.1$, which is constant for all the nodes. Prof. Nicola Perra and prof. Luis E.C. Rocha chose, in their paper, activation probabilities that are extracted from a power-law ($F(p) \sim p^{-1.5}$, with $p \in \interval{0.01}{1}$) \cite{Perra2019}. They did this based on the analyses of real social networks that shows that the propensity of users to start an on-line social interaction follows a heterogeneous distribution \cite{Perra2012}\cite{Perra2019}. However, they also showed that using a constant activation probability $p$ over the entire population yields results that are qualitatively similar (at least in their case) \cite{Perra2019}. \\

@@

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that the other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note here that being stubborn and being extreme doesn't need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Turner}).\\ 
\newline
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of such extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore wouldn't reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

The bounded confidence models, discussed in Section \ref{bounded}, can be easily extended to included some notion of stubbornness. Guillaume Deffuant proposed such an extension where stubbornness is correlated to extremism \cite{Deffuant2006}. The model he proposed goes as follows. Remember that in the bounded confidence models, the agents have an opinion $x_i$ (which is a real number, often between $\interval{0}{1}$ or between $\interval{-1}{1}$) and a threshold $\epsilon_i$, which determines the range of opinions that can influence agent $i$ (agents with opinions between $\interval{x_i - \epsilon_i}{x_i + \epsilon_i}$ are capable of influencing agent $i$). Guillaume Deffuant then introduced a fraction of extremists, $f_e$, which can have either the extreme opinion $-1$ or the extreme opinion $1$ \cite{Deffuant2006}. These extremists are given a threshold equal to $\epsilon_e$ \cite{Deffuant2006}. The other agents are considered to be moderate and can have opinions between $-1$ and $1$. The moderates are given a threshold $\epsilon$, with $\epsilon > \epsilon_e$ \cite{Deffuant2006}. This last condition implicitly denotes the resistance of the extremes towards other opinions in contrast to the more flexible nature of the moderates, which is again based on the idea that people with extreme opinions will be more convinced of that opinion and will be less likely to change it. \\
\newline
Another example of introducing some kind of persistence towards the own opinion of the nodes is the "hardening of positions" that was briefly discussed at the end of Section \ref{classMod}.
% talk about results here??

@@

	\item If $r_i = 1$, the node is completely stubborn and there is no chance of convincing him/here of the other idea

@@

\noindent
\textbf{The threshold model}\\

@@

In the threshold model one introduces a notion of stubbornness by introducing a threshold parameter $T$. This parameter can be either the same for all the nodes ($T$) or can be made individual ($T_i$). The main idea is that nodes can only change their opinion to the other opinion if the fraction of friends with that opinion ($p$) is bigger than $T$ ($p > T$). The idea behind this threshold model is that people are a bit hesitant to change their opinion, but if they have enough friends with the other opinion, they might be convinced of that opinion as well. The more persistent you are, the more friends you need in order to be convinced of another idea. \\

@@

There are different possibilities to implement this kind of model. One way, which makes use of the majority model, is as follows. First the fraction of friends with opinion A and opinion B, $p_A$ and $p_B$ respectively, are determined; then the following rules are used to update the opinion

@@

% Talk about probabilistic implementation? + talk about fact that nothing happens for T < 0.5???

@@

\chapter{Network models: Derivations}
\section{The Erd\H{o}s-R\'{e}nyi model}\label{ER}
\subsection{Degree distribution}\label{degdis}
The probability that a node is connected to $k$ other nodes and not to the $N-1-k$ others is given by
\begin{equation}
	p^k (1-p)^{N-1-k}
\end{equation}
The number of ways to choose the $k$ nodes among the $N-1$ possible candidates is given by the binomial coefficient
\begin{equation}
	\binom{N-1}{k} = \frac{(N-1)!}{k!(N-1-k)!}
\end{equation}
and the probability of being connected to exactly $k$ other nodes becomes
\begin{equation}\label{degDistRan}
	p_k = \binom{N-1}{k}p^k (1-p)^{N-1-k}
\end{equation}
We thus derived the degree distribution of a random network.
